\documentclass{research4cacm}

%%% Note to copy editors: the following fixes are needed
%%% in order to get decent output in the temporary version
%%% for the submission
\makeatletter
\renewcommand{\secfnt}{\fontfamily{ptm}\fontsize{12}{14}\bfseries}
\renewcommand{\subsecfnt}{\fontfamily{ptm}\fontsize{11}{13}\itshape}
\def\section{%
    \@startsection{section}{1}{\z@}{-10\p@ \@plus -4\p@ \@minus -2\p@}% GM
    {14\p@}{\secfnt\@ucheadtrue}%
}

\def\subsection{%
    \@startsection{subsection}{2}{\z@}{-8\p@ \@plus -2\p@ \@minus -\p@}
    {14\p@}{\secfnt}%
}
\def\subsubsection{%
    \@startsection{subsubsection}{3}{\z@}{-8\p@ \@plus -2\p@ \@minus -\p@}%
    {14\p@}{\subsecfnt}%
}
\expandafter\def\expandafter\abstract\expandafter{\abstract\vspace{-\baselineskip}}
\makeatother
%%% end of fixes

%\usepackage[T1]{fontenc}
%\usepackage{lmodern}
%\usepackage[a-2b]{pdfx}
%\usepackage{amsmath}
%\usepackage{amsfonts}
%%\usepackage[utf8]{inputenc}
\usepackage{array}
%\usepackage{stmaryrd}
\usepackage{tikz}
\usepackage{comment}
\usepackage{xspace}
\usepackage{listofitems}
%\usepackage{graphicx}
\usepackage[final]{listings}
\usepackage{enumitem}
%%\usepackage{amsthm}
\usepackage{ifthen}
\usepackage{url}
%\renewcommand{\floatpagefraction}{.8}

% space saving tricks
\newif\ifstreamexamples
\streamexamplestrue
\newif\ifzsetexamples
\zsetexamplestrue
%%\newtheoremstyle{note} % name
%%{2pt} % Space above
%%{2pt} % Space below
%%{}    % Body font
%%{}    % Indent amount
%%{\bfseries} % Theorem head font
%%{:}   % Punctuation after theorem head
%%{.5em}% ⟨Space after theorem head
%%{}    % Theorem head spec (can be left empty, meaning ‘normal’
%
%\numberwithin{equation}{section}
%
%\graphicspath{ {.} }
\lstset{basicstyle=\ttfamily}
\lstset{language=Java,
  commentstyle=\color{brown},
  keywordstyle=\color{blue},
  stringstyle=\color{red}}

\lstdefinelanguage{ddlog}{
  language=Java, % we base it on Java, just for comments
  morekeywords={input, output, typedef, relation, typedef, bool, not,
    string, bit, extern, function, var, for, match, skip, in, integer, % not really in DDlog
    Aggregate, FlatMap},
  deletestring=[b]{'}
}
%\hypersetup{
%  colorlinks   = true,    % Colours links instead of ugly boxes
%  urlcolor     = blue,    % Colour for external hyperlinks
%  linkcolor    = blue,    % Colour of internal links
%  citecolor    = red      % Colour of citations
%}
%\hypersetup{final}
%
\usetikzlibrary{shapes, arrows.meta, positioning,
  decorations.pathreplacing, matrix}
\tikzstyle{block}=[draw,rectangle]
\tikzstyle{every node}=[font=\small]
%\theoremstyle{note}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{example}[theorem]{Example}
\newtheorem{algorithm}[theorem]{Algorithm}

\permission{Copyright is held by the owner/author(s). Publication
  rights licensed to the VLDB Endowment. This is a simplified version
  of the paper entitled \dbsp: Automatic Incremental View Maintenance
  for Rich Query Languages, published in PVLDB, Vol. 16, Issue 7, ISSN
  2150-8097~\cite{budiu-vldb23}.  DOI:
  https://doi.org/10.14778/3587136.3587137}
\copyrightetc{}

\begin{document}

% Used when a term is first defined.  Adds the term to the index.
\newcommand{\defined}[1]{\textbf{#1}\index{}}
\newcommand{\zr}{$\Z$-set\xspace}
\newcommand{\zrs}{$\Z$-sets\xspace} % plural
\newcommand{\means}[1]{\ensuremath{\llbracket #1 \rrbracket}}
\newcommand{\code}[1]{\mbox{\texttt{#1}}}
\newcommand{\Z}{\mathbb{Z}}  % integers
\newcommand{\N}{\mathbb{N}}  % naturals
\newcommand{\B}{\mathbb{B}}  % Booleans
\newcommand{\R}{\mathbb{R}}  % reals
% stream with elements of a given type
\newcommand{\stream}[1]{\ensuremath{\mathcal{S}_{#1}}}
% finite stream with elements of a given type (zero almost everywhere)
\newcommand{\streamf}[1]{\ensuremath{\overline{\mathcal{S}_{#1}}}}
\newcommand{\zm}{\ensuremath{z^{-1}}} % stream delay operator
\ifthenelse{\equal{1}{1}}{ % allows switching to mathit/mathcal
\newcommand{\I}{\mathcal{I}}  % stream integration
\newcommand{\D}{\mathcal{D}}  % stream derivative
}{
\newcommand{\I}{\mathit{I}}  % stream integration
\newcommand{\D}{\mathit{D}}  % stream derivative
}
\newcommand{\inc}[1]{{#1}^{\Delta}}
\newcommand{\distinct}{\mathit{distinct}}  % distinct operator
% set with elements of given type
\newcommand{\secref}[1]{Section~\ref{#1}}  % reference to a section
\newcommand{\refsec}[1]{\secref{#1}}
\newcommand{\set}[1]{\mathit{set}_{#1}}
\newcommand{\id}{\ensuremath{\mathit{id}}} % identity function
\newcommand{\isset}{\mbox{isset}}
\newcommand{\ispositive}{\mbox{ispositive}}
\newcommand{\defn}{\stackrel{\textrm{\scriptsize def}}{=}}
\newcommand{\map}{\mbox{map}}
\newcommand{\fix}[2]{\mbox{fix}\,#1.#2}
\newcommand{\lift}[1]{{\uparrow}#1}
\newcommand{\rew}{\ensuremath{\mapsto}} % rewriting
\newcommand{\birew}{\ensuremath{\mapsfrom\!\mapsto}} % bidirectional rewriting
\newcommand{\pair}[2]{\ensuremath{\langle #1,#2 \rangle}} % pairing
\newcommand{\norm}[1]{\| #1 \|} % norm; requires math mode
%\newcommand{\zpp}[1]{\mbox{zpp}(#1)}
\newcommand{\makeset}{\ensuremath{\mbox{makeset}}}
\newcommand{\sv}[1]{ % simple stream value, supplied as a space-separated list of 5 values
\setsepchar{ }
\readlist\arg{#1}
\begin{tabular}{|c|c|c|c|c|} \hline
  % notice that this is in reverse
  \!\ensuremath{\cdots}\! &
  \!\arg[4]\! &
  \!\arg[3]\! &
  \!\arg[2]\! &
  \!\arg[1]\!
  % \arg[5] &
  \\ \hline
\end{tabular}
}
\newcommand{\dbsp}{DBSP\xspace}

\numberofauthors{5}
\author{
  \alignauthor Mihai Budiu \\
  \affaddr{Feldera} \\
  \email{mbudiu@feldera.com} \\
  \alignauthor Tej Chajed \\
  \affaddr{Univ. of Wisconsin-Madison} \\
  \email{chajed@wisc.edu} \\
  \alignauthor Frank McSherry \\
  \affaddr{Materialize Inc.} \\
  \email{mcsherry@materialize.com} \\
  \and
  \alignauthor Leonid Ryzhyk \\
  \affaddr{Feldera} \\
  \email{leonid@feldera.com} \\
  \alignauthor Val Tannen \\
  \affaddr{University of Pennsylvania} \\
  \email{val@seas.upenn.edu} \\
}

\title{\dbsp: Incremental Computation on Streams and Its Applications
  to Databases} \maketitle

\begin{abstract}
  We describe \dbsp, a framework for incremental computation.
  Incremental computations repeatedly evaluate a function on some
  input values that are ``changing''.  The goal of an efficient
  implementation is to ``reuse'' previously computed results.
  Ideally, when presented with a new change to the input, an
  incremental computation should only perform work proportional to the
  size of the changes of the input, rather than to the size of the
  entire dataset.

  In databases ``incremental computation'' is known as Incremental
  View Maintenance (IVM); IVM has long been a central problem of
  database theory and practice.  Many solutions have been proposed for
  restricted classes of computation or of changes, but we are seeking
  a general solution.

  We start by defining incremental computations as computations on
  data streams, i.e., sequences of data values, by borrowing ideas
  from Digital Signal Processing.

  Using these tools, we give a general solution to the incremental
  computation problem in 4 steps: (1) we describe a simple but
  expressive language called \dbsp for describing computations over
  data streams; (2) we give a new mathematical definition of
  incremental computation for \dbsp programs; (3) we give a general
  algorithm for converting any \dbsp program into an incremental
  program.  The algorithm reduces the problem of incrementalizing a
  complex query to the problem of incrementalizing the primitive
  operations that compose the query. Finally, (4) we show that
  practical database query languages, such as SQL and Datalog, can be
  directly implemented on top of \dbsp, using primitives with
  efficient incremental implementations.  As a consequence, we obtain
  a general recipe for efficient IVM for essentially arbitrary queries
  written in all these languages.
\end{abstract}

\input{intro}
\input{streams}
\input{relational}
\input{recursion}
\input{nested}
%\input{extensions}
\input{implementation}
%\input{related}
\input{conclusions}

\newpage

\bibliographystyle{abbrv}
\bibliography{main}

\end{document}
